Bumble bee communities are structured by a spatiotemporal mosaic of competitive

### 2020-05-13 manuscript.Rmd
functionalized breakpoints

---
title: Bumble bee communities are structured by a spatiotemporal mosaic of competitive
  processes
author:
- Douglas B. Sponsler
- Katharina Kallnik
- Fabrice Requier
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
output:
  word_document:
    reference_docx: docx_template.docx
bibliography: refs.bib

---

@@

### Load site data
site_data <- read_csv("./processed_data/site_data.csv") %>%
  #dplyr::select(-c(slope.calc, slope.est, elev.min, elev.max, temp.mean)) %>% # drop these variables
  mutate(elev.class = factor(elev.class, levels = c("oben", "mitte", "unten")),
         elev.class2 = factor(elev.class2, levels = c("high", "mid", "low"))) # turn elav.class into an ordered factor

@@

*Colony phase*
```{r}
### Get caste counts
castes <- read_csv("./processed_data/network.csv") %>%
  group_by(site, date, year, caste) %>%
  summarize(caste.count = n()) %>%
  spread(caste, caste.count) %>%
  replace(is.na(.), 0) %>%
  mutate(q.prop = queen/(queen+worker)) %>%
  left_join(gdd, by = c("site", "date")) %>%
  left_join(site_data, by = c("site")) %>%
  arrange(gdd.cum)

### Logistic regression
glm.qw.2012.high <- glm(q.prop ~ gdd.cum,
                        family = quasibinomial,
                        data = filter(castes, year == "2012",
                                      elev.class2 == "high"))

glm.qw.2012.mid <- glm(q.prop ~ gdd.cum,
                        family = quasibinomial,
                        data = filter(castes, year == "2012",
                                      elev.class2 == "mid"))

glm.qw.2012.low <- glm(q.prop ~ gdd.cum,
                        family = quasibinomial,
                        data = filter(castes, year == "2012",
                                      elev.class2 == "low"))

### Extract gdd corresponding to desired proportion; based on https://stackoverflow.com/questions/32040504/regression-logistic-in-r-finding-x-value-predictor-for-a-particular-y-value
#### Note that this only works for logistic regression!
getX <- function(x, y) {
  (log(y/(1-y)) - coef(x)[1])/coef(x)[2]
}

break1_2012_high <- getX(glm.qw.2012.high, 0.5) %>%
  data.frame() %>%
  select(break1 = 1) %>%
  mutate(elev.class2 = "high",
         year = 2012)

break1_2012_mid <- getX(glm.qw.2012.mid, 0.5)  %>%
  data.frame() %>%
  select(break1 = 1) %>%
  mutate(elev.class2 = "mid",
         year = 2012)

break1_2012_low <- getX(glm.qw.2012.low, 0.5)  %>%
  data.frame() %>%
  select(break1 = 1) %>%
  mutate(elev.class2 = "low",
         year = 2012)

break1_2012 <- bind_rows(break1_2012_high, break1_2012_mid, break1_2012_low) %>%
  mutate(elev.class2 = factor(elev.class2, levels = c("high", "mid", "low")))

# Add break points to castes data frame
castes_2012 <- castes %>%
  filter(year == "2012") %>%
  left_join(break1_2012, by = c("year", "elev.class2"))

# Plot
ggplot(castes_2012, aes(gdd.cum, q.prop)) +
  geom_point() +
  geom_smooth(method = "glm", method.args=list(family="quasibinomial")) +
  geom_vline(aes(xintercept = break1), color = "red") +
  facet_wrap(~elev.class2, ncol = 1)

%>%
  group_by(site, date, year) %>%
  #mutate(caste_prop = caste_count/sum(caste_count)) %>%
  spread(caste, caste_count) %>%
  replace(is.na(.), 0) %>%
  mutate(queen = oldqueen + queen,
         reproductive = male + youngqueen) %>%
  select(-oldqueen, -male, -youngqueen) %>%
  group_by(site, year) %>%
  mutate(worker.cum = cumsum(worker),
         repro.cum = cumsum(reproductive)) %>%
  left_join(gdd, by = c("site", "date")) %>%
  left_join(site_data, by = c("site")) %>%
  select(site, date, year, queen, worker, reproductive, gdd.cum, elev.class, elev.class2, elev.mean) %>%
  mutate(q.w = queen/worker)

  gather(caste, count, -c(site, date, year, gdd.cum, elev.class, elev.class2, elev.mean)) %>%
  group_by(site, date) %>%
  mutate(prop = count/sum(count))

glm_queen_2012 <- glm(prop ~ gdd.cum, 
                      family = quasibinomial(),
                      data = filter(castes, year == "2012" & caste == "queen" & elev.class2 == "mid"))

beta_queen_2012 <- betareg(prop ~ gdd.cum, 
                      data = filter(castes, year == "2012" & caste == "queen" & elev.class2 == "mid"))

p <- predict(glm_queen_2012, type = "response")

founding_break <- castes %>%
  spread(caste, count) %>%
  replace(is.na(.), 0) %>%
  select(site, date, year, gdd.cum, queen, worker, reproductive) %>% 
  filter(worker > queen & queen > 0) %>%
  group_by(site, year) %>%
  filter(date == min(date)) %>%
  select(site, year, queen, worker, reproductive, founding.break = gdd.cum)

repro_break <- castes %>%
  spread(caste, count) %>%
  replace(is.na(.), 0) %>%
  select(site, date, year, gdd.cum, queen, worker, reproductive) %>% 
  filter(reproductive / (reproductive + worker) >= 0.1) %>%
  group_by(site, year) %>%
  filter(date == min(date)) %>%
  select(site, year, queen, worker, reproductive, repro.break = gdd.cum)

phases <- castes %>%
  full_join(founding_break, by = c("site", "year")) %>%
  full_join(repro_break, by = c("site", "year"))

# ggplot(phase, aes(gdd.cum, prop, color = caste)) +
#   geom_point(alpha = 0.25) +
#   geom_smooth() +
#   theme_gray(10) +
#   facet_grid(elev.class2~year, scales = "free") +
#   xlab("Cumulative growing degree days") +
#   ylab("Proportional abundance")
# 
# ggsave("./output/gdd_phase.pdf", width = 6, height = 4)

ggplot(filter(castes, year == "2012"), aes(gdd.cum, q.w)) +
  geom_point() +
  geom_smooth() +
  #geom_vline(aes(xintercept = founding.break), linetype = "dotted") +
  facet_wrap(~elev.class2)

ggplot(filter(phases, year == "2011"), aes(gdd.cum, prop, color = caste)) +
  geom_point() +
  geom_smooth() +
  #geom_vline(aes(xintercept = founding.break), linetype = "dotted") +
  facet_wrap(~elev.class2, scales = "free")

ggplot(filter(phases, year == "2010"), aes(gdd.cum, prop, color = caste)) +
  geom_point() +
  geom_smooth() +
  #geom_vline(aes(xintercept = founding.break), linetype = "dotted") +
  facet_wrap(~elev.class2, scales = "free")

```

@@

>>>>

---
title: 
author:
- Douglas B. Sponsler
- Katharina Kallnik
- Fabrice Requier
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
output:
  word_document:
    reference_docx: docx_template.docx
bibliography: refs.bib
---

Conversation with Ingolf
- Add opening paragraph setting context of clinate change and community ecology
- Look for divergent population trajectories of species as evidence of competition with temporal lag on effects
- Where does network analysis fit in? A legit question -- no need to jump right to fancy network metrics if simpler approaches work. But so far, none of my empirical predictions involve network metrics. 
- Ingolf had a REALLY nice title -- ask him to send it to me.
-  Ingolf said that he does not think that competition from syrphids at high elevation would be significant, but competition from honey bees at low elevation could be
- Switch Fabrice and Kalli's order
- Cite Ingolf's recent papers showing the effect of pulses of mass-flowering crop resource on bumble bee health
- He liked my colony phase figures

@@

### Load site data
site_data <- read_csv("./processed_data/site_data.csv") %>%
  dplyr::select(-c(slope.calc, slope.est, elev.min, elev.max, temp.mean)) %>% # drop these variables
  mutate(elev.class = factor(elev.class, levels = c("oben", "mitte", "unten")),
         elev.class2 = factor(elev.class2, levels = c("high", "mid", "low"))) # turn elav.class into an ordered factor

@@

*Colony phase*
```{r}
### Get caste counts
castes <- read_csv("./processed_data/network.csv") %>%
  group_by(site, date, year, caste) %>%
  summarize(caste.count = n()) %>%
  spread(caste, caste.count) %>%
  replace(is.na(.), 0) %>%
  mutate(queen = oldqueen + queen,
         reproductive = male + youngqueen,
         q.prop = queen/(queen+worker+reproductive),
         r.prop = reproductive/(queen+worker+reproductive)) %>%
  select(-oldqueen, -youngqueen, -male) %>%
  left_join(gdd, by = c("site", "date")) %>%
  left_join(site_data, by = c("site")) %>%
  arrange(gdd.cum)

### Logistic regression for first break point (founding to buildup)
breakpoints <- function(x, year) {

  data <- filter(x, year == !!year)

  # Logistic regression for first break point (founding to buildup)
  glm1.low <- glm(q.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "low"))

  glm1.mid <- glm(q.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "mid"))

  glm1.high <- glm(q.prop ~ gdd.cum,
                   family = quasibinomial,
                   data = filter(data, elev.class2 == "high"))

  # Logistic regression for second break point (buildup to reproductive)
  glm2.low <- glm(r.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "low"))

  glm2.mid <- glm(r.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "mid"))

  glm2.high <- glm(r.prop ~ gdd.cum,
                   family = quasibinomial,
                   data = filter(data, elev.class2 == "high"))

  ### Extract gdd corresponding to desired proportion; based on https://stackoverflow.com/questions/32040504/regression-logistic-in-r-finding-x-value-predictor-for-a-particular-y-value

  getX <- function(x, y) {
    (log(y/(1-y)) - coef(x)[1])/coef(x)[2]
    }

  break1.low <- getX(glm1.low, 0.5) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break1")

  break1.mid <- getX(glm1.mid, 0.5) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "mid",
           breakID = "break1")

  break1.high <- getX(glm1.high, 0.5) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "high",
           breakID = "break1")

  break2.low <- getX(glm2.low, 0.2) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break2")

  break2.mid <- getX(glm2.mid, 0.2) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "mid",
           breakID = "break2")

  break2.high <- getX(glm2.high, 0.2) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "high",
           breakID = "break2")

  # Join into data frame, convert elev.class2 to ordered factor, spread
  breaks <- full_join(break1.low, break1.mid) %>%
    full_join(break1.high) %>%
    full_join(break2.low) %>%
    full_join(break2.mid) %>%
    full_join(break2.high) %>%
    mutate(elev.class2 = factor(elev.class2, levels = c("high", "mid", "low"))) %>%
    spread(breakID, gdd)

  #
  out <- data %>%
    full_join(breaks) %>%
    gather(caste, count, -c(site, date, year, q.prop, r.prop,
                          gdd.cum, elev.class, elev.class2, 
                          management, transect, elev.mean, 
                          lat, lon, break1, break2)) %>%
  group_by(site, date) %>%
  mutate(prop = count/sum(count))
}

breaks_2012 <- breakpoints(castes, 2012)
breaks_2011 <- breakpoints(castes, 2011)
breaks_2010 <- breakpoints(castes, 2010)

# Plot
binomial_smooth <- function(...) {
  geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = FALSE, ...)
}

facet_names <- c(`low` = "Low (641-1000 m)",
                 `mid` = "Mid (1000-1499 m)",
                 `high` = "High (1500-2032 m)")

ggplot(breaks_2012, aes(gdd.cum, prop, color = caste)) +
  geom_point() +
  binomial_smooth(formula = y ~ splines::ns(x, 2)) +
  geom_vline(aes(xintercept = break1), linetype = "dashed") +
  geom_vline(aes(xintercept = break2), linetype = "dashed") +
  facet_wrap(~elev.class2, ncol = 1, labeller = as_labeller(facet_names)) +
  xlab("Cumulative growing degree days") +
  ylab("Proportional abundance")

ggsave("./output/phases2012.pdf")

ggplot(breaks_2011, aes(gdd.cum, prop, color = caste)) +
  geom_point() +
  binomial_smooth(formula = y ~ splines::ns(x, 2)) +
  geom_vline(aes(xintercept = break1), linetype = "dashed") +
  geom_vline(aes(xintercept = break2), linetype = "dashed") +
  facet_wrap(~elev.class2, ncol = 1, labeller = as_labeller(facet_names)) +
  xlab("Cumulative growing degree days") +
  ylab("Proportional abundance")

ggsave("./output/phases2011.pdf")

ggplot(breaks_2010, aes(gdd.cum, prop, color = caste)) +
  geom_point() +
  binomial_smooth(formula = y ~ splines::ns(x, 2)) +
  geom_vline(aes(xintercept = break1), linetype = "dashed") +
  geom_vline(aes(xintercept = break2), linetype = "dashed") +
  facet_wrap(~elev.class2, ncol = 1, labeller = as_labeller(facet_names)) +
  xlab("Cumulative growing degree days") +
  ylab("Proportional abundance")

ggsave("./output/phases2010.pdf")
```

@@

<!-- Bumble bee communities are structured by a spatiotemporal mosaic of competitive -->
<!--   processes -->
